\bf Approximating Likelihood Ratios with Calibrated Discriminative Classifiers

### 2016-02-09 jasa/paper-jasa.tex

\begin{algorithm}[t]
\caption{Learning a parametrized classifier.}\label{alg:training}
\begin{algorithmic}
    \State ${\cal L := \{ \}}$;
    \While{ $\text{size}({\cal L}) < N $ }
        \State Select or draw $\theta_0$, $\theta_1$ from $\Theta_0$, $\Theta_1$;
	    \State Draw $\mathbf{x} \sim p(\mathbf{x}|\theta_0)$;
		\State ${\cal L} := {\cal L} \cup \{ ((\mathbf{x}, \theta_0, \theta_1), y=0) \}$;
		\State Draw $\mathbf{x} \sim p(\mathbf{x}|\theta_1)$;
		\State ${\cal L} := {\cal L} \cup \{ ((\mathbf{x}, \theta_0, \theta_1), y=1) \}$;
    \EndWhile
    \State Learn a single classifier $s(\mathbf{x}; \theta_0, \theta_1)$ from ${\cal L}$.
\end{algorithmic}
\end{algorithm}

@@

Using the method outlined in Section~\ref{sec:approx},
figures~\ref{fig:2}, \ref{fig:3} and \ref{fig:4} show the approximated
ratio when using respectively a linear model, a 2-layer neural network or a random
forest. The blue curve corresponds to the true target log-ratio, the green curve is
the direct approximation $\log \left( \frac{\hat s(\mathbf{x})}{1 - \hat s(\mathbf{x})} \right)$
without calibration, while the red curve is the approximation $\log \frac{\hat p(\hat
s(\mathbf{x}) | \theta_0)}{\hat p(\hat s(\mathbf{x}) | \theta_1)}$ as calibrated
using histograms. Finally, the
cyan curve shows the approximated ratio when decomposing the mixtures, as
described in Section~\ref{sec:mixture}.

\begin{figure}
    \centering
    \begin{subfigure}[b]{0.45\textwidth}
        \includegraphics[width=\textwidth]{figures/fig1.pdf}
        \caption{$p(\mathbf{x}|\theta_0)$ and $p(\mathbf{x}|\theta_1)$}
        \label{fig:1}
    \end{subfigure}
    ~ %add desired spacing between images, e. g. ~, \quad, \qquad, \hfill etc.
      %(or a blank line to force the subfigure onto a new line)
    \begin{subfigure}[b]{0.45\textwidth}
        \includegraphics[width=\textwidth]{figures/fig2-a.pdf}
        \caption{Linear model}
        \label{fig:2}
    \end{subfigure}

@@

    \begin{subfigure}[b]{0.45\textwidth}
        \includegraphics[width=\textwidth]{figures/fig2-b.pdf}
        \caption{Neural network}
        \label{fig:3}
    \end{subfigure}
    ~
    \begin{subfigure}[b]{0.45\textwidth}
        \includegraphics[width=\textwidth]{figures/fig2-c.pdf}
        \caption{Random forest}
        \label{fig:4}
    \end{subfigure}
    \caption{Approximation of the log-likelihood ratio $\log r(\mathbf{x};\theta_0,\theta_1) = \log \frac{p(\mathbf{x}|\theta_0)}{p(\mathbf{x}|\theta_1)}$ with calibrated classifiers.}

\end{figure}

\glnote{Diagnostics.}

\glnote{Extend to illustrate a generalized likelihood ratio test with
parameterized classifier / calibration by assuming that the true value of $g$ is unknown.}

>>>>

\begin{algorithm}[t]
\caption{Learning a parameterized classifier.}\label{alg:training}
\begin{algorithmic}
    \State ${\cal L := \{ \}}$;
    \While{ $\text{size}({\cal L}) < N $ }
        \State Select or draw $\theta_0$, $\theta_1$ from $\Theta_0$, $\Theta_1$;
	    \State Draw $\mathbf{x} \sim p(\mathbf{x}|\theta_0)$;
		\State ${\cal L} := {\cal L} \cup \{ ((\mathbf{x}, \theta_0, \theta_1), y=0) \}$;
		\State Draw $\mathbf{x} \sim p(\mathbf{x}|\theta_1)$;
		\State ${\cal L} := {\cal L} \cup \{ ((\mathbf{x}, \theta_0, \theta_1), y=1) \}$;
    \EndWhile
    \State Learn a single classifier $s(\mathbf{x}; \theta_0, \theta_1)$ from ${\cal L}$.
\end{algorithmic}
\end{algorithm}

@@

Using the method outlined in Section~\ref{sec:approx},
figures~\ref{fig:2a}, \ref{fig:2b} and \ref{fig:2c} show the approximated
ratio when using respectively a linear model, a 2-layer neural network or a random
forest. The blue curve corresponds to the true target log-ratio, the green curve is
the direct approximation $\log \left( \frac{\hat s(\mathbf{x})}{1 - \hat s(\mathbf{x})} \right)$
without calibration, while the red curve is the approximation $\log \frac{\hat p(\hat
s(\mathbf{x}) | \theta_0)}{\hat p(\hat s(\mathbf{x}) | \theta_1)}$ as calibrated
using histograms. Finally, the
cyan curve shows the approximated ratio when decomposing the mixtures, as
described in Section~\ref{sec:mixture}.
Hyper-parameters of the classifiers are all tuned on a finite grid of values
in order to minimize the mean squared error, as estimated by 3-fold cross-validation.

\begin{figure}
    \centering
    \begin{subfigure}[b]{0.45\textwidth}
        \includegraphics[width=\textwidth]{figures/fig1.pdf}
        \caption{$p(\mathbf{x}|\theta_0)$ and $p(\mathbf{x}|\theta_1)$}
        \label{fig:1}
    \end{subfigure}
    ~ %add desired spacing between images, e. g. ~, \quad, \qquad, \hfill etc.
      %(or a blank line to force the subfigure onto a new line)
    \begin{subfigure}[b]{0.45\textwidth}
        \includegraphics[width=\textwidth]{figures/fig2-a.pdf}
        \caption{Linear model}
        \label{fig:2a}
    \end{subfigure}

@@

    \begin{subfigure}[b]{0.45\textwidth}
        \includegraphics[width=\textwidth]{figures/fig2-b.pdf}
        \caption{Neural network}
        \label{fig:2b}
    \end{subfigure}
    ~
    \begin{subfigure}[b]{0.45\textwidth}
        \includegraphics[width=\textwidth]{figures/fig2-c.pdf}
        \caption{Random forest}
        \label{fig:2c}
    \end{subfigure}
    \caption{Approximation of the log-likelihood ratio
             $\log r(\mathbf{x};\theta_0,\theta_1) = \log \frac{p(\mathbf{x}|\theta_0)}{p(\mathbf{x}|\theta_1)}$
             with calibrated classifiers.}
\end{figure}

The first fact to highlight is that, in the finite sample setting, uncalibrated
direct approximations $\log \left( \frac{\hat s(\mathbf{x})}{1 - \hat
s(\mathbf{x})} \right)$, as shown in green, do not yield very accurate estimates
of the ratio, even when using universal approximators such as a neural network
or a random forest. This observation is at the origin of this work, and led us
to show that the ratio $r(\mathbf{x})$ can still be computed exactly by forming
the equivalent statistic
$\frac{p(s(\mathbf{x})|\theta_0)}{p(s(\mathbf{x})|\theta_1)}$, provided the
transformation $s(\mathbf{x})$ is monotonic with $r(\mathbf{x})$. Indeed, when
constructing $\frac{\hat p(\hat s(\mathbf{x})|\theta_0)}{\hat p(\hat
s(\mathbf{x})|\theta_1)}$ as we propose, estimates appear much closer to the
true ratio than their direct counterparts, as illustrated in red. The
approximated ratios show to be accurate \glnote{Would be nice to be quantitative
here.} for all three classifiers, in particular in regions with enough density
(i.e., from $\mathbf{x}=-3$ to $\mathbf{x}=2$) and where the capacity of the
classifier is focused. By contrast, in regions where the densities are low
(i.e., at the boundaries) the approximated ratio show to have high variance
since the underlying classifier cannot make very accurate predictions, because
of the few training samples it is trained on in these parts of the input space.
Accordingly, when leveraging the fact that densities are mixtures and decomposing their
ratio, the capacity of the underlying classifiers can be focused on easier
supervised learning tasks, resulting as expected in even more accurate approximations, as
shown in cyan for all three classifiers.

As shown in the figures, qualitatively different results are obtained depending
the classification algorithm embedded within the approximation of the ratio. In
this case, the neural network-based approximation appears to work best, due to
the capacity of the classifier to model a decision function $s(\mathbf{x})$
continuous and almost monotonic with the ratio. By contrast, the random forest
yields a decision function which is piece-wise constant, therefore non-continuous
and not strictly monotonic with the ratio in the finite sample regime, which
appears to result in more variance in the approximated ratio. Finally, the
linear model produces a decision function that is continuous but only piece-wise
monotonic with the ratio. Due to the symmetry of the ratio around
$\mathbf{x}=1$, this still results in an accurate approximation, but poorer
results should be expected on more complex problems. Similarly, calibrating
$\hat s(\mathbf{x})$ to construct $\hat p(\hat s(\mathbf{x}))$ is also key to
obtain accurate results. While standard histograms with fixed bins have been
used here for illustrative purposes, we anticipate that the embedded density
estimation algorithm should be carefully chosen and tuned, on a case by case
basis.

Let us now consider the composite hypothesis testing setting, by comparing
$H_0: g = 0$ versus the alternative $H_1: g > 0$. That is, we want to evaluate
\begin{equation}
\Lambda({\cal D}; \Theta_0 = \{0\}, \Theta=(0, \frac{1}{2}]) =  \frac{ p({\cal D} | g = 0) }{ \sup_{\theta \in \Theta} p({\cal D} | g=\theta)} \; .
\end{equation}
Following the approach outlined in Section~\ref{sec:generalized-likelihood-ratio},
computing the generalized likelihood ratio statistic requires finding
the maximum likelihood estimator for the denominator, which can be carried out using
approximated likelihood ratios. Reusing the decomposed neural network-based
approximation (or equivalently, using a parameterized classifier) and minimizing the mean negative log-ratio of some observed data
generated for $g=0.05$,
we successfully find as estimator $\hat g=0.049$, as shown in Figure~\ref{fig:3}.
From there, the generalized likelihood ratio can be approximated as in the simple
hypothesis setting.

\begin{figure}
    \centering
    \includegraphics[width=0.5\textwidth]{figures/fig3.pdf}
    \caption{Using approximated likelihood ratios for parameter inference.}
    \label{fig:3}
\end{figure}

\glnote{Smoothly interpolate between histogram bins, in order to get a smooth ratio as well. Worth mentioning in the paper!}
\glnote{Discuss diagnostics?}
\glnote{Better highlight advantages of the method?}
\glnote{Conduct more thorough and convincing experiments?}
